---
title: "Statistical Methods Final Project"
author: "Kyle Urban, James Joyner, Brennon"
date: "`r Sys.Date()`"
output: 
  html_document:
     code_folding: show
     theme: flatly
     highlight: tango
     toc: yes
     number_sections: yes
     df_print: paged
     toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages
```{r warning=F, message=F}
library(tidyverse)

```

```{r}
DATA <- read_csv("Teen_Mental_Health_Dataset.csv")

DATA
```

Qualitative Variables: **gender**, **platform_usage**

Quantitative Variables: **sleep_hours**, **screen_time_before_sleep**
## -------------------------------------------------------------
## Heading 1
## -------------------------------------------------------------

### Section 1.1 Qualitative Data

Qualitative Variable: **gender**

Success: **female**

Population size: N = 1200


------------------------------------------------------------------------------

#### 1. Select one qualitative variable and use one level of the qualitative variable as success and select a sample size of at least 50.

```{r}
# Frequency Table
DATA %>%
  count(gender, name = 'frequency') %>%
  mutate(rel_frequency = frequency/sum(frequency))
```

Qualitative variable = gender

Success = female

Sample size: n = 50

Population proportion: p = 585/1200
```{r}
p <- 585/1200
p
```


------------------------------------------------------------------------------

#### 2. Show the number of successes is from binomial distribution.

Let X = number of females in a random sample of n = 50 teens.

**Checking Binomial Criteria:**

1. The experiment is performed a fixed number of times: n = 50 (fixed).

2. The trials are independent: n <= 0.05N → 50 <= 0.05 * 1200 = 60. Since 50 <= 60, the trials are independent.

3. For each trial, there are two mutually exclusive outcomes: female (success) or male (failure).

4. The probability of success is fixed for each trial: p = 585/1200 = 0.4875.

Since all four criteria are met, X ~ Bin(n = 50, p = 0.4875).

```{r}
# n = sample size
n <- 50

# Check independence: n <= 0.05N
n <= 0.05 * 1200
```

```{r}
# PMF table for X ~ Bin(n = 50, p = 0.4875)
binom_pmf <- tibble(x_female = 0:50,
                    prob_female = dbinom(x = x_female, size = 50, prob = p))
binom_pmf
```

```{r}
# Graph of pmf
binom_pmf %>%
  ggplot(mapping = aes(x = x_female, y = prob_female)) +
  geom_col(width = 0.01, color = "blue") +
  geom_point(size = 1) +
  labs(x = "Number of Females",
       y = "Probability",
       title = "PMF of X ~ Bin(50, 0.4875)")
```


------------------------------------------------------------------------------

#### 3. Find the probability the random variable is more than its mean.

```{r}
# Mean: E(X) = np
mean_x <- n * p
mean_x

# Standard Deviation: SD(X) = sqrt(np(1-p))
sd_x <- sqrt(n * p * (1 - p))
sd_x
```

P(X > 24.375) = P(X >= 25) = 1 - P(X <= 24)

```{r}
# Method 1: Using pbinom
1 - pbinom(q = 24, size = 50, prob = p)
```

```{r}
# Method 2: Summing individual probabilities
dbinom(x = 25:50, size = 50, prob = p) %>% sum()
```


------------------------------------------------------------------------------

#### 4. Find the sampling distribution of the sample proportion of the qualitative variable with a sample size of at least 50.

$\hat{p}$ = sample proportion of females in a sample of n = 50

The sampling distribution of $\hat{p}$:

i) $\mu_{\hat{p}} = p = 0.4875$
```{r}
p
```

ii) $\sigma_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}}$
```{r}
sqrt(p * (1 - p) / n)
```

iii) Approximately normal if:

  a) Simple random sample (assumed)
  
  b) n <= 0.05N: 50 <= 0.05 * 1200 = 60 
  
  c) np(1-p) >= 10
```{r}
n * p * (1 - p)
```

np(1-p) = 12.49 >= 10

Since the requirements hold, $\hat{p}$ is approximately normal.


------------------------------------------------------------------------------
------------------------------------------------------------------------------


### Section 1.2 Quantitative Data

Quantitative Variable: **sleep_hours**


------------------------------------------------------------------------------

#### 1. Select one quantitative variable.

```{r}
DATA %>%
  summarise(mean_sleep = mean(sleep_hours, na.rm = T),
            sd_sleep = sd(sleep_hours, na.rm = T),
            median_sleep = median(sleep_hours, na.rm = T),
            IQR_sleep = IQR(sleep_hours, na.rm = T))
```


------------------------------------------------------------------------------

#### 2. Check if the quantitative variable is from normal distribution.

```{r}
# QQ Plot
DATA %>%
  ggplot(mapping = aes(sample = sleep_hours)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot of Sleep Hours")
```

```{r}
# Histogram
DATA %>%
  ggplot(mapping = aes(x = sleep_hours)) +
  geom_histogram(fill = "blue", color = "black", binwidth = 0.5) +
  labs(y = "Frequency",
       title = "Histogram of Sleep Hours")
```

```{r}
# Boxplot
DATA %>%
  ggplot(mapping = aes(x = sleep_hours)) +
  geom_boxplot(fill = "blue", color = "black") +
  labs(title = "Boxplot of Sleep Hours")
```

The QQ plot shows the points approximately follow the reference line. The histogram shows an approximately symmetric distribution. Based on these plots, sleep_hours is approximately normally distributed.


------------------------------------------------------------------------------

#### 3. Find the sampling distribution of the sample mean of the quantitative variable with sample size of at least 50.

Population parameters (since this section assumes population data):
```{r}
# Population mean
mu_x <- mean(DATA$sleep_hours)
mu_x

# Population standard deviation
sigma_x <- sd(DATA$sleep_hours)
sigma_x
```

Sample size: n = 50 (same as Section 1.1)

The sampling distribution of $\bar{X}$:

i) $\mu_{\bar{X}} = \mu_X$
```{r}
mu_x
```

ii) $\sigma_{\bar{X}} = \sigma_X / \sqrt{n}$
```{r}
sigma_x / sqrt(50)
```

iii) Normal or approximately normal if:

  a) Simple random sample (assumed) YES
  
  b) n <= 0.05N: 50 <= 0.05 * 1200 = 60 YES
  
  c) n >= 30: 50 >= 30  (also sleep_hours ~ approximately Normal) YES

Since the requirements hold, $\bar{X}$ is approximately normal.


------------------------------------------------------------------------------

#### 4. Find a probability of sample mean is greater than a number.

Find $P(\bar{X} > 6.5)$

```{r}
# P(X_bar > 6.5) = 1 - P(X_bar <= 6.5)
1 - pnorm(q = 6.5, mean = mu_x, sd = sigma_x / sqrt(50))
```

## -------------------------------------------------------------
## Heading 2
## -------------------------------------------------------------



## -------------------------------------------------------------
## Heading 3
## -------------------------------------------------------------



## -------------------------------------------------------------
## Heading 4
## -------------------------------------------------------------

#Section 4.0
a) Population is all teenagers
b) Sample is 1200 teenagers

#Section 4.1
a) The variable is daily hours of sleep which is Quantitative
b) Parameter is mean number of hours asleep
c) ?
d) Point Estimate is 6.449417
   Lower = 6.341972 and Upper = 6.556862
```{r}
DATA %>% summarise(DATA_mean = mean(sleep_hours), DATA_sd = sd(sleep_hours))

t.test(x = DATA$sleep_hours, conf.level = 0.99)
```
e) It is recommended by doctors and health professionals that everyone should sleep for 8 hours a night to get a good nights rest. Lack of sleep can lead to multiple mental problems which can affect anyone. Do these results show a average of sleep less then 8 hours a night with a 0.05 significance level?
```{r}
t0 <- (6.449417 - 8)/(1.442677/sqrt(1200))
t0
pt(q = t0, df = 1200 - 1 )
```
There is sufficient evidence at $\alpha$ = 0.05 to conclude that the mean hours of sleep each night for teenagers is less then the recommended 8

#Section 4.2
a) The variable is social interaction level which is qualitative
b) Parameter is proportion of teenagers that use instagram
c)?
d) Point estamte is 0.3425
   Lower
```{r}
DATA %>% count(platform_usage)
411/1200
```


## -------------------------------------------------------------
## Heading 5
## -------------------------------------------------------------



## -------------------------------------------------------------
## Heading 6
## -------------------------------------------------------------



## -------------------------------------------------------------
## Heading 7
## -------------------------------------------------------------


