---
title: "Statistical Methods Final Project"
author: "Kyle Urban, James Joyner, Brennon"
date: "`r Sys.Date()`"
output: 
  html_document:
     code_folding: show
     theme: flatly
     highlight: tango
     toc: yes
     number_sections: yes
     df_print: paged
     toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages
```{r warning=F, message=F}
library(tidyverse)

```

```{r}
DATA <- read_csv("Teen_Mental_Health_Dataset.csv")

DATA
```

Qualitative Variables: **gender**, **platform_usage**

Quantitative Variables: **sleep_hours**, **screen_time_before_sleep**
## -------------------------------------------------------------
## Heading 1
## -------------------------------------------------------------

### Section 1.1 Qualitative Data

Qualitative Variable: **gender**

Success: **female**

Population size: N = 1200


------------------------------------------------------------------------------

#### 1. Select one qualitative variable and use one level of the qualitative variable as success and select a sample size of at least 50.

```{r}
# Frequency Table
DATA %>%
  count(gender, name = 'frequency') %>%
  mutate(rel_frequency = frequency/sum(frequency))
```

Qualitative variable = gender

Success = female

Sample size: n = 50

Population proportion: p = 585/1200
```{r}
p <- 585/1200
p
```


------------------------------------------------------------------------------

#### 2. Show the number of successes is from binomial distribution.

Let X = number of females in a random sample of n = 50 teens.

**Checking Binomial Criteria:**

1. The experiment is performed a fixed number of times: n = 50 (fixed).

2. The trials are independent: n <= 0.05N → 50 <= 0.05 * 1200 = 60. Since 50 <= 60, the trials are independent.

3. For each trial, there are two mutually exclusive outcomes: female (success) or male (failure).

4. The probability of success is fixed for each trial: p = 585/1200 = 0.4875.

Since all four criteria are met, X ~ Bin(n = 50, p = 0.4875).

```{r}
# n = sample size
n <- 50

# Check independence: n <= 0.05N
n <= 0.05 * 1200
```

```{r}
# PMF table for X ~ Bin(n = 50, p = 0.4875)
binom_pmf <- tibble(x_female = 0:50,
                    prob_female = dbinom(x = x_female, size = 50, prob = p))
binom_pmf
```

```{r}
# Graph of pmf
binom_pmf %>%
  ggplot(mapping = aes(x = x_female, y = prob_female)) +
  geom_col(width = 0.01, color = "blue") +
  geom_point(size = 1) +
  labs(x = "Number of Females",
       y = "Probability",
       title = "PMF of X ~ Bin(50, 0.4875)")
```


------------------------------------------------------------------------------

#### 3. Find the probability the random variable is more than its mean.

```{r}
# Mean: E(X) = np
mean_x <- n * p
mean_x

# Standard Deviation: SD(X) = sqrt(np(1-p))
sd_x <- sqrt(n * p * (1 - p))
sd_x
```

P(X > 24.375) = P(X >= 25) = 1 - P(X <= 24)

```{r}
# Method 1: Using pbinom
1 - pbinom(q = 24, size = 50, prob = p)
```

```{r}
# Method 2: Summing individual probabilities
dbinom(x = 25:50, size = 50, prob = p) %>% sum()
```


------------------------------------------------------------------------------

#### 4. Find the sampling distribution of the sample proportion of the qualitative variable with a sample size of at least 50.

$\hat{p}$ = sample proportion of females in a sample of n = 50

The sampling distribution of $\hat{p}$:

i) $\mu_{\hat{p}} = p = 0.4875$
```{r}
p
```

ii) $\sigma_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}}$
```{r}
sqrt(p * (1 - p) / n)
```

iii) Approximately normal if:

  a) Simple random sample (assumed)
  
  b) n <= 0.05N: 50 <= 0.05 * 1200 = 60 
  
  c) np(1-p) >= 10
```{r}
n * p * (1 - p)
```

np(1-p) = 12.49 >= 10

Since the requirements hold, $\hat{p}$ is approximately normal.


------------------------------------------------------------------------------
------------------------------------------------------------------------------


### Section 1.2 Quantitative Data

Quantitative Variable: **sleep_hours**


------------------------------------------------------------------------------

#### 1. Select one quantitative variable.

```{r}
DATA %>%
  summarise(mean_sleep = mean(sleep_hours, na.rm = T),
            sd_sleep = sd(sleep_hours, na.rm = T),
            median_sleep = median(sleep_hours, na.rm = T),
            IQR_sleep = IQR(sleep_hours, na.rm = T))
```


------------------------------------------------------------------------------

#### 2. Check if the quantitative variable is from normal distribution.

```{r}
# QQ Plot
DATA %>%
  ggplot(mapping = aes(sample = sleep_hours)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot of Sleep Hours")
```

```{r}
# Histogram
DATA %>%
  ggplot(mapping = aes(x = sleep_hours)) +
  geom_histogram(fill = "blue", color = "black", binwidth = 0.5) +
  labs(y = "Frequency",
       title = "Histogram of Sleep Hours")
```

```{r}
# Boxplot
DATA %>%
  ggplot(mapping = aes(x = sleep_hours)) +
  geom_boxplot(fill = "blue", color = "black") +
  labs(title = "Boxplot of Sleep Hours")
```

The QQ plot shows the points approximately follow the reference line. The histogram shows an approximately symmetric distribution. Based on these plots, sleep_hours is approximately normally distributed.


------------------------------------------------------------------------------

#### 3. Find the sampling distribution of the sample mean of the quantitative variable with sample size of at least 50.

Population parameters (since this section assumes population data):
```{r}
# Population mean
mu_x <- mean(DATA$sleep_hours)
mu_x

# Population standard deviation
sigma_x <- sd(DATA$sleep_hours)
sigma_x
```

Sample size: n = 50 (same as Section 1.1)

The sampling distribution of $\bar{X}$:

i) $\mu_{\bar{X}} = \mu_X$
```{r}
mu_x
```

ii) $\sigma_{\bar{X}} = \sigma_X / \sqrt{n}$
```{r}
sigma_x / sqrt(50)
```

iii) Normal or approximately normal if:

  a) Simple random sample (assumed) YES
  
  b) n <= 0.05N: 50 <= 0.05 * 1200 = 60 YES
  
  c) n >= 30: 50 >= 30  (also sleep_hours ~ approximately Normal) YES

Since the requirements hold, $\bar{X}$ is approximately normal.


------------------------------------------------------------------------------

#### 4. Find a probability of sample mean is greater than a number.

Find $P(\bar{X} > 6.5)$

```{r}
# P(X_bar > 6.5) = 1 - P(X_bar <= 6.5)
1 - pnorm(q = 6.5, mean = mu_x, sd = sigma_x / sqrt(50))
```

## -------------------------------------------------------------
## Heading 2
## -------------------------------------------------------------
### 2.1 Research Objectives

**Main Research Question:** Does screen time before sleep on social media platforms affect sleep hours, and does this relationship differ by gender?

------------------------------------------------------------------------------

#### 1. One sample confidence interval for the population proportion from the variable selected in Section 1.1.

Construct a 95% confidence interval for the proportion of teens who are female.

This establishes the demographic makeup of the population. Before comparing sleep patterns between genders, we need to understand the gender distribution and how confident we are in that estimate.

------------------------------------------------------------------------------

#### 2. One sample confidence interval for the population mean from the variable selected in Section 1.2.

Construct a 95% confidence interval for the mean sleep hours of teens.

This gives us a range for the true average sleep hours across all teens. Before investigating whether screen time is hurting sleep, we need a baseline estimate of how much sleep teens are actually getting.

------------------------------------------------------------------------------

#### 3. Formulate a research question to conduct a one-sample proportion hypothesis test based on the variable from Section 1.1.

**Research Question:** Is the proportion of female teens in the population different from 50%?

$H_0: p = 0.50$

$H_A: p \neq 0.50$

This tests whether the gender split in the population is roughly even or skewed. If the dataset is heavily skewed toward one gender, that matters when we interpret the two-sample comparison of sleep hours later.

------------------------------------------------------------------------------

#### 4. Formulate a research question to conduct a one sample t test hypothesis test based on the variable in Section 1.2.

**Research Question:** Do teens get less than the recommended 8 hours of sleep?

$H_0: \mu = 8$

$H_A: \mu < 8$

The CDC recommends 8-10 hours of sleep for teens. Testing whether the mean sleep hours is significantly less than 8 directly addresses the core concern of our research question. If teens are not getting enough sleep, it motivates investigating screen time as a potential cause.

------------------------------------------------------------------------------

#### 5. Create a research question to do two sample mean (independent t test) hypothesis tests.

**Research Question:** Is there a difference in mean sleep hours between male and female teens?

$H_0: \mu_1 - \mu_2 = 0$ (no difference in mean sleep hours between males and females)

$H_A: \mu_1 - \mu_2 \neq 0$ (there is a difference in mean sleep hours between males and females)

This directly answers the second half of the main research question: "does this relationship differ by gender?" By comparing mean sleep hours between males and females, we can determine if one gender is sleeping significantly less than the other.

-------------------------------------------------------------------------------

#### 6. Use two quantitative variables to show if there is a linear relationship between them.

**Research Question:** Is there a linear relationship between screen time before sleep and sleep hours?

This is the core of the entire research question. By fitting a regression model with screen_time_before_sleep as the explanatory variable and sleep_hours as the response variable, we can determine whether more screen time before bed predicts less sleep, how strong that relationship is, and by how much sleep changes per unit increase in screen time.


## -------------------------------------------------------------
## Heading 3
## -------------------------------------------------------------


What to do: Use the variables from your research question in Introduction section

a)Find the distribution of the two quantitative variables: center, spread and shape. 

#Finding Shape
is it symmetric or skewed? Is it bimodal or unimodal? 
#Quantitative Variable 1: **sleep_hours**
```{r}


DATA %>%
  summarise(
    mean_val = mean(sleep_hours, na.rm = TRUE),
    median_val = median(sleep_hours, na.rm = TRUE),
    diff = mean_val - median_val 
  )



DATA %>%
  summarise(
    Shape = "Approximately Symmetric",
    Center_Mean = mean(sleep_hours, na.rm = TRUE),
    Spread_SD = sd(sleep_hours, na.rm = TRUE),
  )

```
Since the difference between the mean value & the median value is near zero (diff = -0.05058333)
We can say that the distribution of sleep hours is approximately normal



```{r}

#Quantitative Variable 2: **screen_time_before_sleep**


```

#Finding Center
Mean (symmetric) or median (skewed)?
```{r}

```

#Finding spread
If symmetric, find the standard deviation
If skewed, find the IQR.
```{r}


```

b)Create bar graph and relative frequency table for the two categorical variables. 



## -------------------------------------------------------------
## Heading 4
## -------------------------------------------------------------

#Section 4.0
a) Population is all teenagers
b) Sample is 1200 teenagers

#Section 4.1
a) The variable is daily hours of sleep which is Quantitative
b) Parameter is mean number of hours asleep
c) Independent
d) Point Estimate is 6.449417
   Lower = 6.341972 and Upper = 6.556862
```{r}
DATA %>% summarise(DATA_mean = mean(sleep_hours), DATA_sd = sd(sleep_hours))

t.test(x = DATA$sleep_hours, conf.level = 0.99)
```
e) It is recommended by doctors and health professionals that everyone should sleep for 8 hours a night to get a good nights rest. Lack of sleep can lead to multiple mental problems which can affect anyone. Do these results show a average of sleep less then 8 hours a night with a 0.05 significance level?
```{r}
t0 <- (6.449417 - 8)/(1.442677/sqrt(1200))
t0
pt(q = t0, df = 1200 - 1 )
```
There is sufficient evidence at $\alpha$ = 0.05 to conclude that the mean hours of sleep each night for teenagers is less then the recommended 8

#Section 4.2
a) The variable is gender which is Qualitative
b) Parameter is the proportions of females in the study
c) Independent
d) Point Estimate is 48.75% 
   Lower = 43.95% and Upper = 53.55%
```{r}
DATA %>% count(gender)
585/1200

p_hat <- 0.4875
s_hat <- sqrt(0.4875*(1-0.4875)/1200)
z_alpha_2 <- 1 - qnorm((1-0.98)/2, mean = 0, sd = 1)

#Lower Bound
p_hat - z_alpha_2 * s_hat

#Upper Bound
p_hat + z_alpha_2 * s_hat
```
e) The gender background of subjects can show trends just as much as any other qualitative variables. At 0.5 significance level, test the hypothesis that the proportion of female teenagers is less then 50%.
```{r}
prop.test(x = 585, n = 1200, p = 0.50, alternative = "less", correct = F)
```
Since p-value = 0.1932 < 0.05 = $\aplha$, we can reject $H_0$. There is sufficient evidence at $\aplha$ = 0.5 to conclude that the proportion of teenagers that were female was less the 50%

## -------------------------------------------------------------
## Heading 5
## -------------------------------------------------------------


Quantitative Variables: **sleep_hours**
Test female vs male sleep hours

#Section 5.1

We want to test if the time asleep for males is largely different time spent asleep for females. We will be testing this at the 95% confidence level with $\mu_1 - \mu_2$ representing the difference from male sleeping to female sleeping.
a) Population 1 = all male teenagers
   Population 2 = all female teenagers
b) Sample 1 = 615 male teenagers
   Sample 2 = 585 female teenagers
c) Time asleep and gender, Time Alseep is Quantitiative while Gender is Qualitative
d) Parameter 1 = $\mu_1$ = mean time asleep for males
   Parameter 2 = $\mu_2$ = mean time asleep for females
e) They are both independent
```{r}
DATA_male <- DATA %>% filter(gender == "male")
DATA_female <- DATA %>% filter(gender == "female")

t.test(x = DATA_male$sleep_hours, y = DATA_female$sleep_hours, alternative = "less", conf.level = 0.95)
```
Since 0 is in the 95% confidence interval, we fail to reject the NULL hypothesis. There is insufficient evinced using 95% confidence interval to conclude that the mean difference between male and female teenager sleeping hours is less then 0.


## -------------------------------------------------------------
## Heading 6
## -------------------------------------------------------------


Use two quantitative variables to show if there is a linear relationship between them.


a) What are the variables? What type of variables are they? 

Quantitative Variable 1: **sleep_hours**
Quantitative Variable 2: **screen_time_before_sleep**


b) Run a hypothesis test to test your hypothesis.


#Not sure if it wants me to test the original hypothesis from section 2.1 or the new hypothesis made specifically for this section? The instructions for the project do not make this clear. As such, might be safe to test both. 

Hypothesis 1: Is there a linear relationship between the screen_time_before_sleep and the number of sleep hours gotten? 
$H_0: \beta_1 = 0$
$H_A: \beta_1 \neq 0$

```{r}


model = lm(screen_time_before_sleep ~ sleep_hours, data = DATA)

#coef(model)
summary(model)
```

P-value: 0.7232

Decision: The P-value of 0.7232 is greater than alpha level of $\alpha = 0.05$. Also, the Adjusted R-squared is near zero, showing only a little of variance is a result of the dependent variable. Thus we fail to reject the null hypothesis

Conclusion: Given the P-value of 0.7232 is greater than alpha level $\alpha = 0.05$, we fail to reject the null hypothesis as there is insufficient evidence that there is a linear relationship between the quantitative variables of screen time before sleep and sleep hours obtained. 

#Does the linear relationship vary by gender? 
Hypothesis 2: Is there a linear relationship between the screen_time_before_sleep and the number of sleep hours gotten for females only? 

$H_0: \beta_f = 0$
$H_A: \beta_f \neq 0$


```{r}

#Does it vary by gender? 
fem <- DATA %>% filter(gender == "female")
model_fem = lm(screen_time_before_sleep ~ sleep_hours, data = fem)
summary(model_fem)


```

P-value: 0.3316

Conclusion: P-value of 0.3316 is greater than alpha level $\alpha = 0.05$. Therefore, we fail to reject the null hypothesis as there is insufficient evidence that there is a linear relationship between the quantitative variables of screen time before sleep and sleep hours obtained for only females. 



Hypothesis 3: Is there a linear relationship between the screen_time_before_sleep and the number of sleep hours gotten for males only? 

$H_0: \beta_m = 0$
$H_A: \beta_m \neq 0$



```{r}
male <- DATA %>% filter(gender == "male")
model_male = lm(screen_time_before_sleep ~ sleep_hours, data = male)
summary(model_male)
```


P-value: 0.1857

Conclusion: The P-value of 0.1857 is greater than alpha level $\alpha = 0.05$. Thus, we fail to reject the null hypothesis as there is insufficient evidence that there is a linear relationship between the quantitative variables of screen time before sleep and sleep hours obtained for males only.



Section 6 Disucssion: 

Based on the fact that we failed to reject the null hypothesis that there is a linear relationship between screen time before sleep and sleep hours for both genders, just females, and just males. It is safe to say that there is not a linear relationship between our two quanitative variables. 



## -------------------------------------------------------------
## Heading 7
## -------------------------------------------------------------


