---
title: "Statistical Methods Final Project"
author: "Kyle Urban, James Joyner, Brennon"
date: "`r Sys.Date()`"
output: 
  html_document:
     code_folding: show
     theme: flatly
     highlight: tango
     toc: yes
     number_sections: yes
     df_print: paged
     toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages
```{r warning=F, message=F}
library(tidyverse)

```

```{r}
DATA <- read_csv("Teen_Mental_Health_Dataset.csv")

DATA
```

## -------------------------------------------------------------
## Heading 1
## -------------------------------------------------------------



## -------------------------------------------------------------
## Heading 2
## -------------------------------------------------------------



## -------------------------------------------------------------
## Heading 3
## -------------------------------------------------------------



## -------------------------------------------------------------
## Heading 4
## -------------------------------------------------------------

#Section 4.0
a) Population is all teenagers
b) Sample is 1200 teenagers

#Section 4.1
a) The variable is daily hours of sleep which is Quantitative
b) Parameter is mean number of hours asleep
c) ?
d) Point Estimate is 6.449417
   Lower = 6.341972 and Upper = 6.556862
```{r}
DATA %>% summarise(DATA_mean = mean(sleep_hours), DATA_sd = sd(sleep_hours))

t.test(x = DATA$sleep_hours, conf.level = 0.99)
```
e) It is recommended by doctors and health professionals that everyone should sleep for 8 hours a night to get a good nights rest. Lack of sleep can lead to multiple mental problems which can affect anyone. Do these results show a average of sleep less then 8 hours a night with a 0.05 significance level?
```{r}
t0 <- (6.449417 - 8)/(1.442677/sqrt(1200))
t0
pt(q = t0, df = 1200 - 1 )
```
There is sufficient evidence at $\alpha$ = 0.05 to conclude that the mean hours of sleep each night for teenagers is less then the recommended 8

#Section 4.2
a) The variable is social interaction level which is qualitative
b) Parameter is proportion of teenagers that use instagram
c)?
d) Point estamte is 0.3425
   Lower
```{r}
DATA %>% count(platform_usage)
411/1200
```


## -------------------------------------------------------------
## Heading 5
## -------------------------------------------------------------



## -------------------------------------------------------------
## Heading 6
## -------------------------------------------------------------



## -------------------------------------------------------------
## Heading 7
## -------------------------------------------------------------


